shorts: An R Package for Modeling Short Sprints

### 2021-04-04 paper.tex
PeerJ version of the manuscript

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Mladen Jovanović\\University of Belgrade \And Jason D. Vescovi\\University of Toronto
}
\title{\pkg{shorts}: An R Package for Modeling Short Sprints}

\Plainauthor{Mladen Jovanović, Jason D. Vescovi}

\Abstract{
Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \pkg{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \pkg{shorts} package to help alleviate errors within the subsequent parameter outcomes.
}

\Keywords{sprinting, sports, modeling}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Mladen Jovanović\\
    University of Belgrade\\
    Faculty of Sport and Physical Education\\
Belgrade, Serbia\\
  E-mail: \email{coach.mladen.jovanovic@gmail.com}\\

      Jason D. Vescovi\\
    University of Toronto\\
    Faculty of Kinesiology and Physical Education\\
Graduate School of Exercise Science\\
Toronto, ON Canada\\
  E-mail: \email{vescovij@gmail.com}\\

% Pandoc citation processing

% Pandoc header
\usepackage[htt]{hyphenat} \usepackage{booktabs} \usepackage{microtype}

@@

\usepackage{amsmath}

@@

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 s) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

In the current paper, we will provide an explanation of one commonly used mathematical equation to model short sprints, modeling applications using the \pkg{shorts} package, issues that can arise during measurement and estimation, and potential solutions to those problems.

@@

Short sprints have been modeled using the mono-exponential equation \eqref{eq:velocity-time} originally proposed by \citet{doi:10.1098/rspb.1927.0035}, and more recently popularized by \citet{clarkNFLCombine40Yard2017}, and \citet{samozinoSimpleMethodMeasuring2016}. Equation \eqref{eq:velocity-time} represents function for instantaneous horizontal velocity \(v\) given the time \(t\) and two model parameters:

@@

The parameters of the equation \eqref{eq:velocity-time} are \textit{maximum sprinting speed} (MSS; expressed in \(ms^{-1}\)) and \textit{relative acceleration} (TAU). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; \textit{maximal acceleration}, expressed in \(ms^{-2}\)) \eqref{eq:maximal-acceleration}.

@@

\begin{CodeChunk}
\begin{table}

@@

\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

@@

\caption[Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds]{Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds.}\label{fig:four-athletes-kinematics}
\end{figure}
\end{CodeChunk}

@@

\caption[Acceleration-Velocity profile of four athletes with different MSS and MAC parameters]{Acceleration-Velocity profile of four athletes with different MSS and MAC parameters.}\label{fig:four-athletes-profile}
\end{figure}
\end{CodeChunk}

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \pkg{shorts} package using non-linear least squares regression implemented in the \texttt{nls()} function in the \pkg{base} R \citep{R-base} and \texttt{nlme()} function in the \pkg{nlme} package \citep{R-nlme} for the mixed-effect models.

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function \citep{R-LambertW}. MSS and TAU parameters are estimated using \texttt{model\_using\_splits()} function:

\begin{CodeChunk}
\begin{CodeInput}
R> require(shorts)
R> 
R> split_distance <- c(10, 20, 30, 40)
R> 
R> split_time <- c(2.17, 3.43, 4.60, 5.73)
R> 
R> m1 <- model_using_splits(
+   distance = split_distance,
+   time = split_time
+ )
R> 
R> m1
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00249   1.00000  -0.00178   0.00265   0.00265   0.00176   0.00157 
#>      MAPE 
#>   0.04742
\end{CodeOutput}
\end{CodeChunk}

@@

Besides providing \emph{residual standard error} (RSE), \pkg{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls()} function:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m1)
\end{CodeInput}
\begin{CodeOutput}
#> 
#> Formula: corrected_time ~ TAU * I(LambertW::W(-exp(1)^(-distance/(MSS * 
#>     TAU) - 1))) + distance/MSS + TAU
#> 
#> Parameters:
#>     Estimate Std. Error t value Pr(>|t|)    
#> MSS   9.0121     0.0155     581 0.000003 ***
#> TAU   1.3083     0.0066     198 0.000025 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 0.00249 on 2 degrees of freedom
#> 
#> Number of iterations to convergence: 4 
#> Achieved convergence tolerance: 0.00000311
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> coef(m1)
\end{CodeInput}
\begin{CodeOutput}
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00
\end{CodeOutput}
\end{CodeChunk}

To create a simple plot of the model, use S3 \texttt{plot()} method, which returns \pkg{ggplot2} \citep{R-ggplot2} object:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(m1) + theme_bw(8)
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Predict time at distance
R> predict_time_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 2.17 3.43 4.60 5.73
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Predict acceleration at time
R> predict_acceleration_at_time(
+   time = c(0, 1, 2, 3, 4, 5, 6),
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.8884 3.2075 1.4935 0.6954 0.3238 0.1508 0.0702
\end{CodeOutput}
\end{CodeChunk}

@@

To estimate force production at distance or time (using \texttt{predict\_force\_at\_distance()} and \texttt{predict\_force\_at\_time()} functions), as well as power production (using \texttt{predict\_power\_at\_distance()} and \texttt{predict\_power\_at\_time()} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{get\_air\_resistance()} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \citet{arsacModelingEnergetics100m2002}, \citet{samozinoSimpleMethodMeasuring2016}, and \citet{vaningenschenauCanCyclePower1991} for more information):

\begin{CodeChunk}
\begin{CodeInput}
R> get_air_resistance(
+   velocity = 5,
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # To calculate horizontal force produced
R> predict_force_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 119.0  58.6  36.9  28.2
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # To calculate power produced
R> predict_power_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 868 490 323 251
\end{CodeOutput}
\end{CodeChunk}

>>>>

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\providecommand{\tightlist}{
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{lipsum}
\usepackage[unicode=true]{hyperref}
\usepackage{longtable}

% Pandoc syntax highlighting
% See https://github.com/rstudio/rticles/issues/182
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc Header
\usepackage{lipsum} \usepackage[htt]{hyphenat} \usepackage{booktabs} \usepackage{microtype}

@@

\title{shorts: An R Package for Modeling Short Sprints}

\author[1]{Mladen Jovanović}

\corrauthor[1]{Mladen Jovanović}{\href{mailto:coach.mladen.jovanovic@gmail.com}{\nolinkurl{coach.mladen.jovanovic@gmail.com}}}
\author[2]{Jason D. Vescovi}

\affil[1]{Faculty of Sport and Physical Education, University of Belgrade, Serbia}
\affil[2]{Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

%

\begin{abstract}
Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \textbf{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \textbf{shorts} package to help alleviate errors within the subsequent parameter outcomes.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

\flushbottom
\maketitle
\thispagestyle{empty}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 s) of maximal efforts (\protect\hyperlink{ref-mangineSpeedForcePower2014}{Mangine et al. 2014}), whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) (\protect\hyperlink{ref-ward-smithEnergyConversionStrategies2001}{Ward-Smith 2001}), but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)(\protect\hyperlink{ref-brownAssessmentLinearSprinting2004}{Brown, Vescovi, and Vanheest 2004}). Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). Radar and laser technology are frequently used laboratory-grade methods (\protect\hyperlink{ref-buchheitMechanicalDeterminantsAcceleration2014}{Buchheit et al. 2014}; \protect\hyperlink{ref-edwardsSprintAccelerationCharacteristics2020}{Edwards et al. 2020}; \protect\hyperlink{ref-jimenez-reyesRelationshipVerticalHorizontal2018}{Jiménez-Reyes et al. 2018}; \protect\hyperlink{ref-marcote-pequenoAssociationForceVelocity2019}{Marcote-Pequeño et al. 2019}) but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \textbf{shorts} package (\protect\hyperlink{ref-R-shorts}{Jovanovic 2020}), written in the R language (\protect\hyperlink{ref-R-base}{R Core Team 2020}), represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}).

In the current paper, we will provide an explanation of one commonly used mathematical equation to model short sprints, modeling applications using the \textbf{shorts} package, issues that can arise during measurement and estimation, and potential solutions to those problems.

@@

Short sprints have been modeled using the mono-exponential equation \eqref{eq:velocity-time} originally proposed by \protect\hyperlink{ref-doi:10.1098ux2frspb.1927.0035}{Furusawa, Hill, and Parkinson} (\protect\hyperlink{ref-doi:10.1098ux2frspb.1927.0035}{1927}), and more recently popularized by \protect\hyperlink{ref-clarkNFLCombine40Yard2017}{Clark et al.} (\protect\hyperlink{ref-clarkNFLCombine40Yard2017}{2017}), and \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al.} (\protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{2016}). Equation \eqref{eq:velocity-time} represents function for instantaneous horizontal velocity \(v\) given the time \(t\) and two model parameters:

@@

The parameters of the equation \eqref{eq:velocity-time} are \emph{maximum sprinting speed} (MSS; expressed in \(ms^{-1}\)) and \emph{relative acceleration} (TAU). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; \emph{maximal acceleration}, expressed in \(ms^{-2}\)) \eqref{eq:maximal-acceleration}.

@@

\begin{table}

@@

@@

\begin{figure}

@@

\caption{Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds.}\label{fig:four-athletes-kinematics}
\end{figure}

@@

\caption{Acceleration-Velocity profile of four athletes with different MSS and MAC parameters.}\label{fig:four-athletes-profile}
\end{figure}

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \textbf{shorts} package using non-linear least squares regression implemented in the \texttt{nls()} function in the \textbf{base R} (\protect\hyperlink{ref-R-base}{R Core Team 2020}) and \texttt{nlme()} function in the \textbf{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2020}) for the mixed-effect models.

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function (\protect\hyperlink{ref-R-LambertW}{Goerg 2020}). MSS and TAU parameters are estimated using \texttt{model\_using\_splits()} function:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(shorts)}

\NormalTok{split\_distance }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{20}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{40}\NormalTok{)}

\NormalTok{split\_time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{2.17}\NormalTok{, }\FloatTok{3.43}\NormalTok{, }\FloatTok{4.60}\NormalTok{, }\FloatTok{5.73}\NormalTok{)}

\NormalTok{m1 }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_splits}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{time =}\NormalTok{ split\_time}
\NormalTok{)}

\NormalTok{m1}
\CommentTok{\#\textgreater{} Estimated model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.01                1.31                6.89 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.52                0.00                0.00 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}   0.00249   1.00000  {-}0.00178   0.00265   0.00265   0.00176 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}   0.00157   0.04742}
\end{Highlighting}
\end{Shaded}

@@

Besides providing \emph{residual standard error} (RSE), \textbf{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls()} function:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{summary}\NormalTok{(m1)}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Formula: corrected\_time \textasciitilde{} TAU * I(LambertW::W({-}exp(1)\^{}({-}distance/(MSS * }
\CommentTok{\#\textgreater{}     TAU) {-} 1))) + distance/MSS + TAU}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Parameters:}
\CommentTok{\#\textgreater{}     Estimate Std. Error t value Pr(\textgreater{}|t|)    }
\CommentTok{\#\textgreater{} MSS   9.0121     0.0155     581 0.000003 ***}
\CommentTok{\#\textgreater{} TAU   1.3083     0.0066     198 0.000025 ***}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Residual standard error: 0.00249 on 2 degrees of freedom}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Number of iterations to convergence: 4 }
\CommentTok{\#\textgreater{} Achieved convergence tolerance: 0.00000311}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{coef}\NormalTok{(m1)}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.01                1.31                6.89 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.52                0.00                0.00}
\end{Highlighting}
\end{Shaded}

To create a simple plot of the model, use S3 \texttt{plot()} method, which returns \textbf{ggplot2} (\protect\hyperlink{ref-R-ggplot2}{Wickham, Chang, et al. 2020}) object:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(m1) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{(}\DecValTok{8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Predict time at distance}
\FunctionTok{predict\_time\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 2.17 3.43 4.60 5.73}

\CommentTok{\# Predict acceleration at time}
\FunctionTok{predict\_acceleration\_at\_time}\NormalTok{(}
  \AttributeTok{time =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{6}\NormalTok{),}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 6.8884 3.2075 1.4935 0.6954 0.3238 0.1508 0.0702}
\end{Highlighting}
\end{Shaded}

@@

To estimate force production at distance or time (using \texttt{predict\_force\_at\_distance()} and \texttt{predict\_force\_at\_time()} functions), as well as power production (using \texttt{predict\_power\_at\_distance()} and \texttt{predict\_power\_at\_time()} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{get\_air\_resistance()} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \protect\hyperlink{ref-arsacModelingEnergetics100m2002}{Arsac and Locatelli} (\protect\hyperlink{ref-arsacModelingEnergetics100m2002}{2002}), \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al.} (\protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{2016}), and \protect\hyperlink{ref-vaningenschenauCanCyclePower1991}{van Ingen Schenau, Jacobs, and de Koning} (\protect\hyperlink{ref-vaningenschenauCanCyclePower1991}{1991}) for more information):

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{get\_air\_resistance}\NormalTok{(}
  \AttributeTok{velocity =} \DecValTok{5}\NormalTok{,}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 6.1}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# To calculate horizontal force produced}
\FunctionTok{predict\_force\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 119.0  58.6  36.9  28.2}
